---
title: "Haiti Neonatal Sepsis Plotting Notebook"
#output:
#  pdf_document: default
#  html_notebook: default
---

```{r loadlib, echo=T, results='hide', message=F, warning=F}
library(fpp3)
library(magrittr)
library(dplyr)
library(aweek)
library(openxlsx)
library(surveillance)
library(grid)
library(gridExtra)
```


We filter the time series and take only data between July 2014 to January 2018.

```{r, message=F}


# Load data frame, data frame with updated microbiology and data frame with updated sepsis information
df <- read.xlsx('data/data.xlsx')
df2 <- read.xlsx('data/data_micro_update.xlsx')
dfs <- read.xlsx('data/data_sepsis_update.xlsx')


# Transform epi week into dates
df <- df %>% mutate(date= week2date(epiwk))
df2 <- df2 %>% mutate(date= week2date(epiwk))
dfs <- dfs%>% mutate(date= week2date(epiwk))

# Rename antibiotic columns
df <- df %>% mutate(ab1 = ampi_total+genta_total, 
                    ab2= cefta_total,
                    ab3 = amik_total,
                    ab4 =  mero_total)

# Only use data between July 2014 and the end of 2017
df <- df %>% filter(between(date, as.Date('2014-07-01'), as.Date('2017-12-31') ) ) %>% replace(is.na(.),0)
df2 <- df2 %>% filter(between(date, as.Date('2014-07-01'), as.Date('2017-12-31') ) ) %>% replace(is.na(.),0)
dfs <-  dfs %>% filter(between(date, as.Date('2014-07-01'), as.Date('2017-12-31') ) ) %>% replace(is.na(.),0)


# Define the time series objects and the moving averages of relevant quantities
TSs <- df %>% as_tsibble()
TSss <- dfs %>% as_tsibble()
TSs2 <- df2 %>% as_tsibble()


TSs <- TSs  %>% mutate(
    ma_neo_adm = slider::slide_dbl(neo_adm, mean,
                .before = 2, .after = 2, .complete = TRUE)
  )

TSs <- TSs  %>% mutate(
    ma_mat_normal_preg = slider::slide_dbl(mat_normal_preg, mean,
                .before = 2, .after = 2, .complete = TRUE)
  )

TSs <- TSs  %>% mutate(
    ma_mat_exit = slider::slide_dbl(mat_exit, mean,
                .before = 2, .after = 2, .complete = TRUE)
  )

TSs <- TSs  %>% mutate(
    ma_neo_exit = slider::slide_dbl(neo_exit, mean,
                .before = 2, .after = 2, .complete = TRUE)
  )

TSs <- TSs  %>% mutate(
    ma_neo_death = slider::slide_dbl(neo_death, mean,
                .before = 2, .after = 2, .complete = TRUE)
  )

TSs <- TSs  %>% mutate(
    ab0 = slider::slide_dbl(ampi_total, mean,
                .before = 2, .after = 2, .complete = TRUE))

TSs <- TSs  %>% mutate(
    ab1 = slider::slide_dbl(genta_total, mean,
                .before = 2, .after = 2, .complete = TRUE))

TSs <- TSs  %>% mutate(
    ab2 = slider::slide_dbl(mero_total, mean,
                .before = 2, .after = 2, .complete = TRUE))

TSs <- TSs  %>% mutate(
    ab3 = slider::slide_dbl(cefta_total, mean,
                .before = 2, .after = 2, .complete = TRUE))

TSs <- TSs  %>% mutate(
    ab4 = slider::slide_dbl(amik_total, mean,
                .before = 2, .after = 2, .complete = TRUE))

TSs2 <- TSs2  %>% mutate(
    ma_gn_bc_sample = slider::slide_dbl(gn_bc_sample, mean,
                .before = 2, .after = 2, .complete = TRUE))

TSss <- TSss  %>% mutate(
    ma_sepsis_los_adm = slider::slide_dbl(sepsis_los_adm, mean,
                .before = 2, .after = 2, .complete = TRUE)
  )

TSss$ma_neo_adm <-  TSs$ma_neo_adm
TSss$neo_adm <-  TSs$neo_adm


```


These are the plots for Figure 1 and Figure 2 in the paper, a preview is shown.

```{r, message=F, warning=F}


plot_neo_adm <- TSs %>% 
  autoplot(ma_neo_adm, size = 1.2, pal=(viridisLite::inferno(begin=0.2, end=0.9, n=9)) ) +
  geom_line(aes(y = neo_adm), size=0.3) +
  scale_x_date(date_labels="%Y",date_breaks  ="1 year", limits=c(as.Date('2013/12/15'), as.Date('2018/01/31'))) +
  ylab("Neonatal admissions (n)")+
  xlab("")+ theme(legend.position = "none", axis.title.x=element_blank(),
        axis.text.x=element_blank(),
        axis.ticks.x=element_blank())
plot_neo_adm


plot_maternal_exit <- TSs %>% 
  autoplot(ma_mat_exit, size = 1.2, pal=(viridisLite::inferno(begin=0.2, end=0.9, n=9)) ) +
  geom_line(aes(y = mat_exit), size=0.3) +
  scale_x_date(date_labels="%Y",date_breaks  ="1 year", limits=c(as.Date('2013/12/15'), as.Date('2018/01/31'))) +
  ylab("Maternal exits (n)")+
  xlab("Date") 
plot_maternal_exit

plot_maternal_normal <- TSs %>% 
  autoplot(ma_mat_normal_preg, size = 1.2, pal=(viridisLite::inferno(begin=0.2, end=0.9, n=9)) ) +
  geom_line(aes(y = mat_normal_preg), size=0.3) +
  scale_x_date(date_labels="%Y",date_breaks  ="1 year", limits=c(as.Date('2013/12/15'), as.Date('2018/01/31'))) +
  ylab("Normal pregnancy admissions (n)")+
  xlab("") + theme(legend.position = c(0.2, 0.8), axis.title.x=element_blank(),
        axis.text.x=element_blank(),
        axis.ticks.x=element_blank())
plot_maternal_normal

plot_neo_mortality <- TSs %>% 
  autoplot(ma_neo_death/ma_neo_exit, size = 1.2, pal=(viridisLite::inferno(begin=0.2, end=0.9, n=9)) ) +
  geom_line(aes(y = neo_death/neo_exit), size=0.3) +
  scale_x_date(date_labels="%Y",date_breaks  ="1 year", limits=c(as.Date('2013/12/15'), as.Date('2018/01/31'))) +
  ylab("Mortality (%)")+
  xlab("Date") 
plot_neo_mortality

plot_sepsis_los <- TSss %>% 
  autoplot(ma_sepsis_los_adm, size = 1.2, pal=(viridisLite::inferno(begin=0.2, end=0.9, n=9))) +
  geom_line(aes(y = sepsis_los_adm), size=0.3) +
  scale_x_date(date_labels="%Y",date_breaks  ="1 year", limits=c(as.Date('2013/12/15'), as.Date('2018/01/31'))) +
  ylab("LO sepsis cases (n)")+
  xlab("") + theme(legend.position = c(0.2, 0.8), axis.title.x=element_blank(),
        axis.text.x=element_blank(),
        axis.ticks.x=element_blank())
plot_sepsis_los

plot_sepsis_los_rate <- TSss %>% 
  autoplot(ma_sepsis_los_adm/ma_neo_adm, size = 1.2, pal=(viridisLite::inferno(begin=0.2, end=0.9, n=9))) +
  geom_line(aes(y = sepsis_los_adm/neo_adm), size=0.3) +
  scale_x_date(date_labels="%Y",date_breaks  ="1 year", limits=c(as.Date('2013/12/15'), as.Date('2018/01/31'))) +
  ylab("LO sepsis rate (%)")+
  xlab("") + theme(legend.position = c(0.2, 0.8), axis.title.x=element_blank(),
        axis.text.x=element_blank(),
        axis.ticks.x=element_blank())
plot_sepsis_los_rate







plot_blood <- TSs2 %>% 
  autoplot(ma_gn_bc_sample, size = 1.2, pal=(viridisLite::inferno(begin=0.2, end=0.9, n=9)) ) +
  geom_line(aes(y = gn_bc_sample), size=0.3) +
  scale_x_date(date_labels="%Y",date_breaks  ="1 year", limits=c(as.Date('2013/12/15'), as.Date('2018/01/31'))) +
  ylab("GNB positive blood cultures (n)")+
  xlab("") + theme(legend.position = "none", axis.title.x=element_blank(),
        axis.text.x=element_blank(),
        axis.ticks.x=element_blank())
plot_blood


plot_ab0 <- TSs %>% 
  autoplot(ab0, size = 1.2, pal=(viridisLite::inferno(begin=0.2, end=0.9, n=9)) ) +
  geom_line(aes(y = ampi_total), size=0.3) +
  scale_x_date(date_labels="%Y",date_breaks  ="year", limits=c(as.Date('2013/12/15'), as.Date('2018/01/31'))) +
  ylab("Ampicillin consumption (n)")+
  xlab("") + theme(legend.position = "none", axis.title.x=element_blank(),
        axis.text.x=element_blank(),
        axis.ticks.x=element_blank())
plot_ab0

plot_ab1 <- TSs %>% 
  autoplot(ab1, size = 1.2, pal=(viridisLite::inferno(begin=0.2, end=0.9, n=9)) ) +
  geom_line(aes(y = genta_total), size=0.3) +
  scale_x_date(date_labels="%Y",date_breaks  ="year", limits=c(as.Date('2013/12/15'), as.Date('2018/01/31'))) +
  ylab("Gentamicin consumption (n)")+
  xlab("") + theme(legend.position = "none", axis.title.x=element_blank(),
        axis.text.x=element_blank(),
        axis.ticks.x=element_blank())
plot_ab1

plot_ab2 <- TSs %>% 
  autoplot(ab2, size = 1.2, pal=(viridisLite::inferno(begin=0.2, end=0.9, n=9)) ) +
  geom_line(aes(y = mero_total), size=0.3) +
  scale_x_date(date_labels="%Y",date_breaks  ="year", limits=c(as.Date('2013/12/15'), as.Date('2018/01/31'))) +
  ylab("Meropenem consumption (n)")+
  xlab("") + theme(legend.position = "none", axis.title.x=element_blank(),
        axis.text.x=element_blank(),
        axis.ticks.x=element_blank())
plot_ab2

plot_ab3 <- TSs %>% 
  autoplot(ab3, size = 1.2, pal=(viridisLite::inferno(begin=0.2, end=0.9, n=9)) ) +
  geom_line(aes(y = cefta_total), size=0.3) +
  scale_x_date(date_labels="%Y",date_breaks  ="year", limits=c(as.Date('2013/12/15'), as.Date('2018/01/31'))) +
  ylab("Ceftazidime consumptio (n)")+
  xlab("Date") 
plot_ab3

plot_ab4 <- TSs %>% 
  autoplot(ab4, size = 1.2, pal=(viridisLite::inferno(begin=0.2, end=0.9, n=9)) ) +
  geom_line(aes(y = amik_total), size=0.3) +
  scale_x_date(date_labels="%Y",date_breaks  ="year", limits=c(as.Date('2013/12/15'), as.Date('2018/01/31'))) +
  ylab("Amikacin consumption (n)")+
  xlab("Date") 
plot_ab4


```


The Figures above are saved in 2 pdfs (page 1 and page 2) here. The code also contains the code to save them to a PDF file. 

```{r, message=F, warning=F}

layout <- rbind(c(1, 2),
                c(3, 4),
                c(5, 6))

#pdf("figures/pdf/figure_1_page1.pdf", 10,8)
#grid.arrange(plot_neo_adm, plot_sepsis_los, plot_maternal_normal, plot_sepsis_los_rate, plot_maternal_exit, plot_neo_mortality, layout_matrix=layout)
#dev.off()

#pdf("figures/pdf/figure_2_page2.pdf", 10,8)
#grid.arrange(plot_blood, plot_ab0, plot_ab1, plot_ab2, plot_ab3, plot_ab4, layout_matrix=layout)
#dev.off()

png("figures/figure_1_tss_page1.png", width=10,height=8, units='in', res=300)
grid.arrange(plot_neo_adm, plot_sepsis_los, plot_maternal_normal, plot_sepsis_los_rate, plot_maternal_exit, plot_neo_mortality, layout_matrix=layout)
dev.off()

png("figures/figure_2_tss_page2.png", width=10,height=8, units='in', res=300)
grid.arrange(plot_blood, plot_ab0, plot_ab1, plot_ab2, plot_ab3, plot_ab4, layout_matrix=layout)
dev.off()

```



Below every combination of alarm/indicator is plotted. As for the previous plots, there is also an option to output everything to a PDF file. 


```{r, message=F, warning=F}

# Redefine the  function stsplot_time1 from the Surveillance package in order to suit our needs.
stsplot_time1_new <- function(
    x, k=1, ylim=NULL, axes=TRUE, xaxis.tickFreq=list("%Q"=atChange),
    xaxis.labelFreq=xaxis.tickFreq, xaxis.labelFormat="%G\n\n%OQ",
    epochsAsDate=x@epochAsDate, xlab="time", ylab="No. infected", main=NULL,
    type="s", lty=c(1,1,2), col=c(NA,1,4), lwd=c(1,1,1),
    outbreak.symbol=list(pch=3, col=3, cex=1, lwd=1),
    alarm.symbol=list(pch=24, col=2, cex=1, lwd=1),
    dx.upperbound=0L, hookFunc=function(){}, .hookFuncInheritance=function() {}, la="a", ld='b',ldd='c',...)
{
  stopifnot(length(k) == 1, is.character(k) || k != 0)

  #Extract slots -- depending on the algorithms: x@control$range
  observed   <- x@observed[,k]
  state      <- x@state[,k]
  alarm      <- x@alarm[,k]
  upperbound <- x@upperbound[,k]
  population <- x@populationFrac[,k]
  binaryTS <- x@multinomialTS

  #Control what axis style is used
  xaxis.dates <- !is.null(xaxis.labelFormat)

  if (binaryTS) {
    observed <- ifelse(population!=0,observed/population,0)
    upperbound <- ifelse(population!=0,upperbound/population,0)
    if (ylab == "No. infected") { ylab <- "Proportion infected" }
  }

  ##### Handle the NULL arguments ######################################
  if (is.null(main) && length(x@control) > 0) {
    #a surveillance algorithm has been run
    action <- switch(class(x), "sts" = "surveillance",
                     "stsNC" = "nowcasting","stsBP" = "backprojection")
    method <- x@control$name
    main <- paste0(action, " using ", method)
  }

  # control where the highest value is
  max <- max(c(observed,upperbound),na.rm=TRUE)

  #if ylim is not specified, give it a default value
  if(is.null(ylim) ){
    ylim <- c(-1/20*max, max)
  }

  # left/right help for constructing the columns
  dx.observed <- 0.5
  upperboundx <- (1:length(upperbound)) - (dx.observed - dx.upperbound)

  #Generate the matrices to plot (values,last value)
  xstuff <- cbind(c(upperboundx,length(observed) + min(1-(dx.observed - dx.upperbound),0.5)))
  ystuff <-cbind(c(upperbound,upperbound[length(observed) ]))

  #Plot the results
  matplot(x=xstuff,y=ystuff,xlab=xlab,ylab=ylab,main=main,ylim=ylim,axes = !(xaxis.dates),type='s', lwd=3,...)

  #This draws the polygons containing the number of counts (sep. by NA)
  i <- rep(1:length(observed),each=5)
  dx <- rep(dx.observed * c(-1,-1,1,1,NA), times=length(observed))
  x.points <- i + dx
  y.points <- as.vector(t(cbind(0, observed, observed, 0, NA)))
  polygon(x.points,y.points,col=col2rgb('slategrey'),border=col[2],lwd=lwd[1])

  #Draw upper bound once more in case the polygons are filled
  
  lines(x=xstuff,y=ystuff,type="s",main=main,lwd=3,...)

  #Draw alarm symbols
  alarmIdx <- which(!is.na(alarm) & (alarm == 1))
  if (length(alarmIdx)>0) {
    matpoints( alarmIdx, rep(-1/40*ylim[2],length(alarmIdx)), pch=alarm.symbol$pch, col=alarm.symbol$col, cex= alarm.symbol$cex, lwd=alarm.symbol$lwd)
  }


  #Label x-axis
  if(xaxis.dates & axes) {
    addFormattedXAxis(x = x, epochsAsDate = epochsAsDate, xaxis.tickFreq = xaxis.tickFreq,
                      xaxis.labelFreq = xaxis.labelFreq, xaxis.labelFormat = xaxis.labelFormat,
                      ...)
  }
  #Label y-axis
  if (axes) {
    axis( side=2 ,...)#cex=cex, cex.axis=cex.axis)
  }
  
  legend( x="topright", 
          legend=c(la, ld, ldd),
          col=c("red", col[2], col[2]), 
          lwd=c(NA, NA, 3), lty=c(NA, NA, lty[1]), pch=c(alarm.symbol$col, 22, NA), pt.bg=c(NA, col2rgb('slategrey'), NA))
  
  #Call hook function for user customized action using the current environment
  environment(hookFunc) <- environment()
  hookFunc()

  #Extra hook functions for inheritance plotting (see e.g. plot function of stsNC objects)
  environment(.hookFuncInheritance) <- environment()
  .hookFuncInheritance()
  invisible()
}

# Load the alarms and indicators
alarms_loaded<- read.csv("data/alarmscomplete_from_py.txt")
indicators_loaded<- read.csv("data/indicatorscomplete_from_py.txt")
indicators <- matrix(c(dfs$sepsis_los_adm/df$neo_adm, df$neo_death/df$neo_exit, df$neo_death/df$neo_exit, dfs$sepsis_los_adm/df$neo_adm, df2$gn_bc_sample), ncol=5)

# Makes sure alarms_loaded has proper column names
names(alarms_loaded) <- c('LOSt', 
       'LOSrt',
       'LOSra',
       'mt',
       'ma',
       'LOSi', 
       'di')


# Define a few labels needed for plotting 

# Indicator label
labels_i <- c('LO sepsis rate indicator', 
       'Mortality indicator', 
       'Future mortality indicator', 
       'Future LO sepsis rate indicator',
       'GNB positive blood culture indicator')

# The units of the indicator
units_i <- c('(%)',
             '(%)',
             '(%)',
             '(%)',
             '(n)')

# Labels for the raw data (in histograms)
labels_hist <- c('LO sepsis rate', 
       'NCU Mortality', 
       'NCU Mortality', 
       'LO sepsis rate',
       'GNB positive blood cultures')

# Labels for the alarms
labels_a <- c('LO sepsis threshold', 
       'LO sepsis rate threshold',
       'LO sepsis rate aberration',
       'Mortality threshold',
       'Mortality aberration',
       'LO sepsis case differential', 
       'Mortality differential')


# The actual plotting

for (j in 1:7){
  for (i in 1:5){
    
    # Group things to plot
    indicator_to_plot <- indicators_loaded[[i]]
    labeli <- labels_i[i]
    labela <- labels_a[j]
    alarm_to_plot <- alarms_loaded[[j]]
    indicator_to_hist <- indicators[,i]
    
    # Start the sts plot and add the proper alarm indicators
    sts_to_plot <- sts(observed = indicator_to_hist, start=c(2014, 28))
    alarms(sts_to_plot) <- matrix(alarm_to_plot)
    upperbound(sts_to_plot) <- matrix(indicator_to_plot)
    
    #pdf(paste0("figures/alarms_pdf/a_", labela ,"_i_",labeli, ".pdf"), width=10,height=5.5)
    #stsplot_time1_new(sts_to_plot, units=1, la=paste0(labela, " alarm"), ld=labels_hist[i], ldd=labeli, ylab=units_i[[i]], xlab='Date')
    #dev.off()
  
    # Creates a figure for each combination. 
    png(paste0("figures/alarms_png/a_", labela ,"_i_",labeli, ".png"), width=10,height=5.5,
        units='in', res=300)
    stsplot_time1_new(sts_to_plot, units=1, la=paste0(labela, " alarm"), ld=labels_hist[i], ldd=labeli, ylab=units_i[[i]], xlab='Date')
    dev.off()
  }
}

```